---
title: "Step 4b - Phyloseq global and local"
subtitle: "Create phyloseq object using global and local output."
authors: "TR Kartzinel, BL Littleford-Colquhoun, HK Hoff, TJ Divoll"
output: html_document
date: "2024-04-25"
html_document:
  df_print: paged
  toc: true
params:
    #don't change these
    user: !r Sys.getenv("LOGNAME")
    data_path: "/oscar/data/tkartzin/projects"
    global_ref_lib_path: "/oscar/data/tkartzin/global_ref_lib"
    local_ref_lib_path: "/oscar/data/tkartzin/local_ref_lib"
 
    #update these for your analysis
    step_3b_analysis_date: "20240529" #YYYYMMDD
    step_3c_analysis_date: "20240530" #YYYYMMDD
    project_code: "test"
    taxa_division: "PLN"
    region_of_interest: "P6"
    metadata_file: "test_phyloseq_metadata.csv"
    percent_match: 1 # value between 0-1; percent match between reference sequence and your sequence
    RRA_threshold: 0.05 # 5%
---

## Create a Phyloseq Object
This notebook will use your data with taxonomy assigned from the latest global reference database as well as the latest project-specific local database to create a phyloseq object for downstream analyses.
```{r setup, include=FALSE}
## Only needed for knitr to render this notebook
knitr::opts_chunk$set(echo = TRUE)

if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, here, tidyr, dplyr, phyloseq, vegan, vegetarian, scales, datawizard)
here::i_am("./Step4a_create_phyloseq_global.Rmd")
```

## Read in data
This code block will use the tabulated file created in Step 3 and a list of samples
```{r}
input_path <- file.path(params$data_path, params$project_code, "merged_runs", sprintf("%s_%s", params$step_3c_analysis_date, params$user))

data_global <- read.csv(file.path(input_path, sprintf("%s_global_alluniq.clean.tag.ann.tab", params$step_3b_analysis_date)), sep="\t", header=TRUE)

data_local <- read.csv(file.path(input_path, sprintf("%s_local_alluniq.clean.tag.ann.tab", params$step_3c_analysis_date)), sep="\t", header=TRUE)

data_samples <- read.csv(params$metadata_file)
```

### Collect so insights on the shape of data
Are there the same number of sequences in your local and global files?
```{r}
if (length(which(data_local$id != data_global$id)) == 0){
  print("Your files have the same number of sequences.")
} else {
  print("There is a mismatch in file lengths. Please check inputs and rerun Step 3.")
}
```

### Rename the default column X to SampleID
```{r}
colnames(data_samples)[colnames(data_samples) == "X"] ="SampleID"
```

### Filter the data to only keep 100% matches in the local set
```{r}
size <- dim(data_local) #number of rows and columns
print(paste0("Number of rows in local data: ", size[1]))
print(paste0("Number of columns in local data: ", size[2]))

best_id_col_local <- find_columns(data_local, starts_with("best_identity"))

data_local <- data_local %>% 
  select(-contains("obiclean_status"))

local_keep <- data_local[which(data_local[,best_id_col_local]==1),]

local_keep_rows <- nrow(local_keep)
print(paste0("Number of local rows kept: ", local_keep_rows))
```


### Filter the data to only keep 100% matches in the global set
Global only keeps perfect matches if they are not already present in the local set of matches
```{r}
size <- dim(data_global) #number of rows and columns
print(paste0("Number of rows in global data: ", size[1]))
print(paste0("Number of columns in global data: ", size[2]))

best_id_col <- find_columns(data_global, starts_with("best_identity"))
best_match_col <- find_columns(data_global, starts_with("best_match"))

data_global <- data_global %>% 
  select(-contains("obiclean_status"))
  
global_keep <- data_global[intersect(which(data_global[,best_id_col]==1), which(data_local[,best_id_col_local]!=1)),]

global_keep$scientific_name <- make.unique(as.character(global_keep$scientific_name))

print(paste0("Number of global rows kept: ", nrow(global_keep)))
```

### Combine the local and global datasets
Combine local and additional global libraries into a database called "keep" for exploratory analyses
```{r}
global_keepKEEP <- global_keep
names(global_keepKEEP) <- names(local_keep)
keep <- rbind(local_keep, global_keepKEEP)
```

### Explore the distribution of sequences
Find out distribution of the sequences you toss both locally and globally (i.e., do some match multiple species perfectly, or anything new in GenBank?)

```{r}
local_global_toss <- data_local[which(data_local$id %in%keep$id==F),] #tells you how many taxa didn't have 100% matches 

print(paste0("The number of sequences not matching at 100%: ", dim(local_global_toss)[1]))
```

### Look at the seqs with highest read counts for matches >98%
```{r}
local_global_toss <- local_global_toss[order(local_global_toss$count,decreasing=T),]
head(local_global_toss)
# may be a couple of taxa we should look at that have 1-2 matches at ~98% match with high read count 
```
### Look at a histogram of %match < 1 for the tossed samples
```{r}
hist(local_global_toss[,best_id_col_local])
```

### Count up 100% matches in each family with the local library
```{r}
local_keep %>% count(family_name, sort = TRUE)
```
### Count up 100% matches in each genera with the local library
```{r}
local_keep %>% count(genus_name, sort = TRUE)
```
### Count up 100% matches in each family with the global library
```{r}
global_keep_families <- global_keep %>% count(family_name, sort = TRUE)
global_keep_families
write.csv(global_keep_families, file.path(input_path, "global_keep_familes.csv"))
```
### Count up 100% matches in each genera with the local library
```{r}
global_keep_genera <- global_keep %>% count(genus_name, sort = TRUE)
global_keep_genera
write.csv(global_keep_genera, file.path(input_path, "global_keep_genera.csv"))
```

## Inspect the 90-99% matches

## Generate summary statistics for publication

## Sample-wise Relative Read Abundance (RRA)

## Disproportionality

## Summarize local and non-trace global

## Summarize sample sizes

## Save outputs and generate Phyloseq object


