---
title: "Step 3c - assign taxonomy w/ local db"
authors: "B.Littleford-Colquhoun & T.Divoll"
output: html_document
date: "2024-04-25"
html_document:
  df_print: paged
  toc: true
params:
    #don't change these
    todaydate: !r (format(Sys.Date(), "%Y%m%d"))
    user: !r Sys.getenv("LOGNAME")
    data_path: "/oscar/data/tkartzin/projects"
    ref_lib_path: "/oscar/data/tkartzin/local_ref_lib"
  
    #update these for your analysis
    reference_lib_project_code: "YNP"
    data_project_code: "test"
    taxa_division: "PLN"
    locus: "trnL"
    region_of_interest: "P6"
    latest_db_date: "20240430" #look in `ref_lib_path/project_code/locus/region_of_interest` for the most recent local reference library you want to use
---

This notebook can be used to run the OBITools 1.2.12 pipeline and serve as a tutorial. Run through each code chunk and inspect the outputs. We assume that you are running this notebook on the same day that you ran Step 3a.

```{r include=FALSE}
## Only needed for knitr to render this notebook
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, here, tidyr, dplyr, readxl, stringr, spgs, filesstrings)
here::i_am("./Step3c_taxonomy_assignment_local.Rmd")
```

#### Set up standard path variables
The following variables are created with the params you input in the YAML block above.
```{r}
output_path <- file.path(params$data_path, params$data_project_code, "merged_runs", params$todaydate, params$user)

cleaned_obitools_fasta <- file.path(output_path, sprintf("%s_alluniq.clean.fasta", params$todaydate))

local_ref_db <- file.path(params$ref_lib_path, params$reference_lib_project_code, params$locus, params$region_of_interest, params$latest_db_date, sprintf("formatdb_%s_%s", params$reference_lib_project_code, params$locus), sprintf("%s_%s_%s", params$reference_lib_project_code, params$locus, params$latest_db_date)) 
#-d in ecotag; ecoPCR taxonomy Database name.

ref_seq <- file.path(params$ref_lib_path, params$reference_lib_project_code, params$locus, params$region_of_interest, params$latest_db_date, sprintf("%s%s_completeDB_%s.fasta", params$reference_lib_project_code, params$region_of_interest, params$latest_db_date)) 
#-R in ecotag; fasta file containing the reference sequences.
```

```{r}
## make environment variables that can be passed to bash code chunks
# set global chunk parameters here
knitr::opts_chunk$set(echo = TRUE)
Sys.setenv(cleaned_obitools_fasta=cleaned_obitools_fasta)
Sys.setenv(local_ref_db=local_ref_db)
Sys.setenv(ref_seq=ref_seq)
Sys.setenv(user=params$user)
Sys.setenv(DATE=params$todaydate)
Sys.setenv(output_path=output_path)
```

#### Assign taxonomy to each sequence using the global reference library
```{bash}
source ~/.bash_profile
conda activate /users/${user}/.conda/envs/obitools-env
mkdir -p $output_path
cd $output_path

ecotag -d ${local_ref_db} -R ${ref_seq} ${cleaned_obitools_fasta} > ${DATE}_local_alluniq.clean.tag.fasta

#Count number of sequences in your file
obicount ${DATE}_local_alluniq.clean.tag.fasta --without-progress-bar
```

Un-useful attributes can be removed from sequences
```{bash}
source ~/.bash_profile
conda activate /users/${user}/.conda/envs/obitools-env
cd $output_path

obiannotate --delete-tag=scientific_name_by_db --delete-tag=obiclean_samplecount \
  --delete-tag=obiclean_count --delete-tag=obiclean_singletoncount \
  --delete-tag=obiclean_cluster --delete-tag=obiclean_internalcount \
  --delete-tag=obiclean_head --delete-tag=taxid_by_db --delete-tag=obiclean_headcount \
  --delete-tag=id_status --delete-tag=rank_by_db --delete-tag=order_name \
  --delete-tag=order ${output_path}/${DATE}_local_alluniq.clean.tag.fasta > \
  ${output_path}/${DATE}_local_alluniq.clean.tag.ann.fasta

#count number of sequences in file (should match above)
obicount -a ${DATE}_local_alluniq.clean.tag.ann.fasta --without-progress-bar
```

### Create output for use in phyloseq
This is your final OBITools output (resembles a large ASV/OTU table) that can be used to create a phyloseq object in the next step
```{bash}
source ~/.bash_profile
conda activate /users/${user}/.conda/envs/obitools-env
cd $output_path

obitab -o ${DATE}_local_alluniq.clean.tag.ann.fasta > ${DATE}_local_alluniq.clean.tag.ann.tab
```
